---
title: "Class 15 Pathway Analysis and Gene Onotology"
author: "Luke Wang"
date: "3/4/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will perfrom more differential expression analysis on a previously published dataset, but this time we will explore furthur and begin to annotate the genes that are differentially expressed.


## Differential Expression Analysis

We first need to perform differential expression analysis on the data set we have before moving forward with more analysis.

```{r}
# loading the DESeq2 library for analysis
library(DESeq2)
```

Now we will load our data files
```{r}
count <- "data/GSE37704_featurecounts.csv"
meta <- "data/GSE37704_metadata.csv"

colData <- read.csv(meta, row.names= 1)
head(colData)
countData <- read.csv(count, row.names = 1)
head(countData)
```

Now we need to do some cleaning up of the data that we just imported because there is an odd length column in the counts data table.

```{r}
countData <- as.matrix(countData[,-1])
head(countData)
```

We also want to remove any genes that have zero count across all the samples.
```{r}
zero <- rowSums(countData)!=0
head(zero)
countData <- countData[zero,]
head(countData)
```

Our data has been cleaned up and now it's time for the DESeq analysis
```{r}
dds <- DESeqDataSetFromMatrix(
  countData = countData,
  colData = colData,
  design = ~condition
)

dds= DESeq(dds)

res <- results(dds, contrast=c("condition","hoxa1_kd", "control_sirna"))
summary(res)
```

We will plot the results on a volcano plot to visualize the fold change.

```{r}
mycols <- rep("gray",nrow(res)) # this will make all the genes gray
mycols [abs(res$log2FoldChange)>2] <- "red" # this will make any genes that has absolute value change of 2 in log2 fold change to be red
inds <- (res$padj<0.01) & (abs(res$log2FoldChange)>2)
mycols[inds] <- "blue"
plot(res$log2FoldChange,-log(res$padj), col=mycols, xlab="Log2(FoldChange", ylab="-Log(P-Value")



```

Now we will add gene annotation to the differential expression analysis data

```{r}
library(AnnotationDbi)
library("org.Hs.eg.db")

columns(org.Hs.eg.db)

res$symbol = mapIds(org.Hs.eg.db,
                    keys=row.names(res), 
                    keytype="ENSEMBL",
                    column="SYMBOL",
                    multiVals="first")

res$entrez = mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="ENTREZID",
                    multiVals="first")

res$name =   mapIds(org.Hs.eg.db,
                    keys=row.names(res),
                    keytype="ENSEMBL",
                    column="GENENAME",
                    multiVals="first")

head(res, 10)
```

```{r}
res <- res[order(res$pvalue),]
write.csv(res, file="deseq_results.csv")
```


## Pathway Analysis

```{r}
library(pathview)
library(gage)
library(gageData)
data("kegg.sets.hs")
data("sigmet.idx.hs")
kegg.sets.hs = kegg.sets.hs[sigmet.idx.hs] # we want to focus on teh metabolic and signaling pathways only
head(kegg.sets.hs)
```

The gage() function require a named vector of fold changes where the names of the values are the Entrez gene IDs. We can extract this information from the annotated Results.

```{r}
foldchanges <- res$log2FoldChange
names(foldchanges) <- res$entrez
head(foldchanges)
```

We will run the gage pathway analysis
```{r}
keggres <- gage(foldchanges, gsets=kegg.sets.hs)
attributes(keggres)
head(keggres$less)
```
Now we can visualize the pathway analysis data with pathview().
```{r}
pathview(gene.data=foldchanges, pathway.id = "hsa04110")
pathview(gene.data = foldchanges, pathway.id = "hsa04110", kegg.native = FALSE)
```

Now let's automate this process and focus on the top 5 upregulated pathways.
```{r}
keggrespathways <- rownames(keggres$greater)[1:5]
keggrespathways
keggresids <- substr(keggrespathways, start=1,stop=8) # this is a cheat way to extract the kegg ids from the strong that was generated
keggresids

pathview(gene.data=foldchanges, pathway.id = keggresids, species = "hsa")
```

## Gene Ontology

```{r}
data("go.sets.hs")
data("go.subs.hs")

gobpset <- go.sets.hs[go.subs.hs$BP] # this restricts the GO to be on biological process

gobpres <- gage(foldchanges,gsets=gobpset, same.dir=T)
gobpres$greater
```

## Reactome Analysis
```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))

write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

