---
title: "Class 10 Structrual Bioinformatics"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PDB Statistics

Analyzing the PDB statistics about percentage of structures solved by X-Ray and Electron Microscopy

```{r}
stats <- read.csv("Data Export Summary.csv", row.names=1) # read csv file and set first column as row name
total <- sum(stats$Total) # get total number of structures
xray <- round(stats$Total[1]/total*100,2) # calculate the percentage of x ray solved structures
em <- round(stats$Total[3]/total*100,2) # calculate the percentage of em solved structures
paste(xray,"% of total structures are solved by X-Ray")
paste(em,"% of total structures are solved by Electron Microscopy")

# Here is an alternative way to get all the percentages
per.by.method <- round(stats$Total/sum(stats$Total)*100,1)
names(per.by.method) <- rownames(stats) # assign row names to each of the vector
paste(per.by.method,"% of total structures are ",rownames(stats), sep = '')
```


Let's look at what proportion of structures are actually protein
```{r}
per.by.protein <- round(sum(stats$Proteins)/sum(stats$Total)*100,1)
paste(per.by.protein,"% of total structures are proteins", sep = '')

```

## Getting started with Bio3D and Bio3D View
```{r}
library(bio3d)
library(bio3d.view)
```

Loading the 1HSG pdb file
```{r}
pdb <- read.pdb("1hsg")
```

Using a developmental tool - bio3d-view to view 1HSG in R
```{r}
view(pdb,"overview",col="sse")
```

## Atom Selection with bio3d

Selecting all alpha carbons
```{r}
# Selecting all the alpha carbon atoms and return their indicies

calpha.inds <- atom.select(pdb,"calpha")
calpha.inds
```

Selecting just the chain A

```{r}
a.inds <- atom.select(pdb,chain="A")
# Selecting C-alphs of chain A
ca.inds <- atom.select(pdb,"calpha",chain="A")
```


Extract the protein only portion of this PDB structure and write it out to a new PDB file

```{r}
pro.inds <- atom.select(pdb,"protein")
pro <- trim.pdb(pdb,inds=pro.inds)
write.pdb(pdb=pro,file="1hsg-no-mk1.pdb")
```



Extract the ligand (i.e drug) and write it to a new PDB file

```{r}
ligand.inds <- atom.select(pdb,"ligand")
mk1 <- trim.pdb(pdb,inds=ligand.inds)
write.pdb(pdb=mk1, file="mk1.pdb")
```

