---
title: "Class 14 Transcriptomics and the analysis of RNA-Seq data"
author: "Luke Wang"
date: "2/27/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# RNA-Seq Analysis

## Import countData and colData

countData contains the information for each gene while colData contain metadata for metadata information.

```{r}
counts <- read.csv("data/airway_scaledcounts.csv", stringsAsFactors = FALSE)
# stringsAsFactors = FALSE prevents the gene names from being convoerted to factors
metadata <- read.csv("data/airway_metadata.csv", stringsAsFactors = FALSE)
```

Now let's examine the data structure of these files. It's very important the column of ID in colData must match the countData row names.

```{r}
head(counts)
head(metadata)
```

How many genes are there in the the counts file
```{r}
nrow(counts)
```

How many experimental conditions are there
```{r}
ncol(counts)
```

Now let's check that the column names of the counts file match the row names of the metadata

```{r}
all(metadata$id == colnames(counts)[-1]) # the -1 will get everything except the first one of the index
# all() will sumarize if all values are true
```


## Differential gene experssion  by visual 

This is solely for demonstration purpose. Don't ever do this for actual differential expression analysis

```{r}
View(metadata)
```

We want to compare the treated vs. control and overall mean of each gene across the 4 different cell types

We want to first get the count data for control
```{r}
control.ids <- metadata$id[metadata[,"dex"]=="control"]
control <- counts[,control.ids]
rownames(control) <- counts$ensgene
head(control)
# this got rid of the gene names but we will get it back now

```


Now we will do it for the treated
```{r}
treated.ids <- metadata$id[metadata[,"dex"]=="treated"]
treated <- counts[,treated.ids]
rownames(treated) <- counts$ensgene
head(treated)
# this got rid of the gene names but we will get it back now
```

We will now calculate the means of each gene between the control and treated
```{r}
control.means <- rowMeans(control)
head(control.means)
treated.means <- rowMeans(treated)
head(treated.means)
```

Now that we are going to combine the control means and treated means together in to a single data frame.

```{r}
meancounts <- data.frame(control.means,treated.means)
head(meancounts)
colSums(meancounts)
```

We want to visualize the mean counts of each gene between the control and treated cell lines
```{r}
plot(meancounts, col=c("red","blue"))
```

The plot sees a lot of points condensed in the corner, we think changin the scale to log scale 

```{r}
plot(meancounts, col=c("red","blue"),log="xy")
```

Now that we see the spread of the means we want to see how the fold change between control and treated. We use log2 becasue the math is easier.
```{r}
meancounts$log2fc <- log2(meancounts$treated.means/meancounts$control.means)
head(meancounts)
```

The negative value means they are below the 45˚ line if we look at the graph. It means when the cell is treated, the gene has a lower value than in control. -Inf means it's trying to log O. We want to filter all the 0 values.

```{r}
zero <- which(meancounts[,1:2]==0, arr.ind=T)
# arr.ind will give the array indicies instead of a messier output
zero.unique <- unique(zero[,1]) 
# we are only caring about the row section of the unique variable because the zero could either be in col 1 or 2
meancounts.correct <- meancounts[-zero.unique,]
head(meancounts.correct)
```

Now that we have cleaned up the data let's set up some threshold to help us furthur characterize the darta. It's common to set the threshold of greater than 2 or less than -2.

```{r}
up.ind <- meancounts.correct$log2fc >2
down.ind <- meancounts.correct$log2fc <(-2)
paste0("Up:",sum(up.ind))
paste0("Down:",sum(down.ind))
```

## Adding Annotation Data

There are couple ways to annotate the gene names, the first method is merging the annotation table that contains all the ENSEMBL gene id with corresponding name to the gene.

```{r}
anno <- read.csv("data/annotables_grch38.csv")
head(anno)
```

We will use the merge() function that merges data together depending on common column frame.

```{r}
counts.anno <- merge(x= meancounts.correct,y= anno, by.x="row.names",by.y="ensgene")
head(counts.anno)
```

An alternative way is to use some Bioconductor annotation packages to help with the annotation.
```{r}
# Bioconductor annotation packages
library("AnnotationDbi")
library("org.Hs.eg.db")
```
```{r}
# Adding the gene symbols
meancounts.correct$symbol <- mapIds(org.Hs.eg.db,
                                    keys=row.names(meancounts.correct),
                                    column="SYMBOL",
                                    keytype="ENSEMBL",
                                    multiVals="first")

# Adding the entrezID
meancounts.correct$entrez <- mapIds(org.Hs.eg.db,
                                    keys=row.names(meancounts.correct),
                                    column="ENTREZID",
                                    keytype="ENSEMBL",
                                    multiVals="first")
# Adding the uniprot ID
meancounts.correct$uniprot <- mapIds(org.Hs.eg.db,
                                    keys=row.names(meancounts.correct),
                                    column="UNIPROT",
                                    keytype="ENSEMBL",
                                    multiVals="first")
head(meancounts.correct)
```

## DESeq2 Analysis
Now we will do the analysis appropriately with DESeq2 library

```{r}
library(DESeq2)
```

DESeq2 uses a special object called DESeqDataSet which is composed from the count data, the metadata and the experimental design. It will also store the data in the object.

```{r}
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = metadata,
                              design= ~dex,
                              tidy= T)
dds
```

Run the DESeq Analysis
```{r}
dds <- DESeq(dds)
```

Getting the results from DESeq Analysis
```{r}
res <- results(dds)
res
summary(res)
```


We can change the p value to 0.05
```{r}
summary(res, alpha=0.05)
```

## Visualizing Data with Volcano Plots

Volcano plots is a common figure to use. X axis shows the fold change and the y axis is a log p value. The higher the value on the y, the more significant they are.

```{r}
plot(res$log2FoldChange, -log(res$padj))

```

Now let's colorize the graph
```{r}

mycols <- rep("gray",nrow(res))
mycols[abs(res$log2FoldChange)>2] = "red"
subsuet.inds <- res$padj<0.01 & abs(res$log2FoldChange)>2
mycols[subsuet.inds] = "blue"
plot(res$log2FoldChange, -log(res$padj), col=mycols)
```

We can also plot this in ggplot
```{r}
library(ggplot2)
ggplot(as.data.frame(res), aes(log2FoldChange, -log10(pvalue), col=mycols)) + 
    geom_point() + 
    ggtitle("Volcano plot")
```

