---
title: "Class 6 Writing Functions in R"
author: "Luke Wang"
date: "1/25/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## File Reading!

Using **data.table()** and related functions to bring data in to R properly.

Insert a code chunk.
You can bring small files into R directly through an URL
```{r}
read.table("https://bioboot.github.io/bggn213_S18/class-material/test1.txt", header= TRUE, sep=",")
```

Assign URL to file1
```{r}
file1 <- "https://bioboot.github.io/bggn213_S18/class-material/test1.txt"
data1 <- read.csv(file1)
data1
```

Reading table with custom seperator ($)
```{r}
file2 <- "https://bioboot.github.io/bggn213_S18/class-material/test2.txt"
data2 <- read.table(file2, header =TRUE, sep="$")
data2
```

If you get a weird data format, try read.table

```{r}
file3 <- "https://bioboot.github.io/bggn213_S18/class-material/test3.txt"
data3 <- read.table(file3)
data3
```

# Writing Function

Writing a function
If in the argument, a variable has a =, then it means it has a default value

## Basic Addition Function
```{r}
add <- function(x, y=1) {
# Sum the input x and y 
  x+y
}
```

Let's use the **add()** function
```{r}
add(1)
```

```{r}
add(1,5)
```

```{r}
add(c(1,2,3,4))
```

```{r}
add(c(1,2,3,4),c(1,2,3,4))
```
What happens with the unexpected?
```{r}
# add(1,2,3)

# Unused argument
```

```{r}
# add(x=1,y="b")
# Non numeric argument to binary operator
```

## Writing More Complex Function

### Rules of Writing a Function
Simplify & Reduce Calculation Duplication

### Rescale Function
You need a *name*, *arguments*, and *body*
```{r}
rescale <- function(x) {
  rng <- range(x)
  (x - rng[1]) / (rng[2] - rng[1])
}
```

Testing with Small Vector
```{r}
rescale(1:10)
```

### Exceptions in Functions

```{r}
rescale (c(1,2,NA,3,10))
```

Diagnosing Unexpected Result
```{r}
z <- c(1,2,NA,3,10)
# range(x)
```

Dealing with NA
```{r}
# Range() has a arugment for na.rm
rescale2 <- function(x) {
  rng <- range(x, na.rm=TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
```

```{r}
rescale2 (c(1,2,NA,3,10))
```

Some more complex function
```{r}
rescale3 <- function(x, na.rm=TRUE, plot=FALSE) {
   if(na.rm) {
     rng <-range(x, na.rm=na.rm)
   } else {
     rng <-range(x)
   }
   print("Hello")
   answer <- (x - rng[1]) / (rng[2] - rng[1])
   return(answer)
   print("is it me you are looking for?")
   if(plot) {
      plot(answer, typ="b", lwd=4)
     print("Please don't ever sing again")
}
   print("I can see it in ...")
   
   # return() will end the function and return whatever variable in the return
}
```

```{r}
rescale3(1:10, plot=T)
```

## Anallysis of Protein Drug Interaction with Bio3D

### Using Bio3D
```{r}
# Load bio3D
library(bio3d)
s1 <- read.pdb("4AKE")  # kinase with drug
s2 <- read.pdb("1AKE")  # kinase no drug
s3 <- read.pdb("1E4Y")  # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s3, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
hc <- hclust( dist( rbind(s1.b, s2.b, s3.b) ) )
plot(hc)
```
Examing what **read.pdb** does
```{r}
library(bio3d)
pdb <- read.pdb("1hbs")
pdb
```

