---
title: "PDB Structure Analysis"
author: "Luke Wang"
date: "1/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Here we will create a function **pdb_analysis()** to analyze .pdb files using the Bio3D package. 

The function in its current state will take any vector of PDB accession numbers, and analyze all the C-alpha atoms on the "A" chain of the protein. The results will be returned in individual plots of each residue plotted against its B-factor as well as secondary structure information on the top and bottom margin of the plot. 

## Inputs for pdb_analysis()
There are only two arguments for the function. 

First is the object that contains pdb accession numbers. This argument has no default value.

Second is the *plot* argument. This argument will set whether to plot the b factor values. The default is set at TRUE. If the argument is set to false, there will be no output from this function. 

## Output for pdb_analysis()
If the arugment *plot* is set to true, then the function will generate plots of B factors of CA atoms on the A chain for each protein that is being analyzed. If argument *plot* is set to false, there is no output from the function.

## Original Code
The objective of this excercise is to reduce the complexity of the original code listed below

```{r}
## Original Code
# s1 <- read.pdb("4AKE")  # kinase with drug
# s2 <- read.pdb("1AKE")  # kinase no drug
# s3 <- read.pdb("1E4Y")  # kinase with drug
# s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
# s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
# s3.chainA <- trim.pdb(s3, chain="A", elety="CA")
# s1.b <- s1.chainA$atom$b
# s2.b <- s2.chainA$atom$b
# s3.b <- s3.chainA$atom$b
# plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
# plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
# plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

## New Function!

Here we will create a function based off of the original code listed above that will reduce the complexity of this analysis.
```{r}
# Load bio3D packages for functions to analyze PDB files
library(bio3d)

# Establishing the pdb_analysis function with 2 arguments
pdb_analysis <- function(protein, plot=TRUE){
  for (x in protein){
    protein1 <- read.pdb(x)
    pro1.chainA <- trim.pdb(protein1, chain ="A", elety = "CA")
    pro1.b <- pro1.chainA$atom$b
    if (plot == TRUE){
      plotb3(pro1.b, sse=pro1.chainA, typ="l", ylab="Bfactor")
    }
  }
}

```


## Some Examples

```{r}
# Example with a vector of 3 proteins, with default plot setting
pdb_analysis(c("4AKE", "1AKE", "1E4Y"))

# Example with a string of single protein, with default plot setting
pdb_analysis("4AKE")

# Example with a vector of 3 proteins but no plots
pdb_analysis(c("4AKE", "1AKE", "1E4Y"), plot = FALSE)
# No useful output is produced from the previous example. The protein that was analyzed will not be passed outside of the function for furthur use.
```

