---
title: "Class 7 R Functions and Packages"
author: "Luke Wang"
date: "1/30/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Revisiting Functions

Revisiting the first example function we created last week - rescale()

Load in the rescale function with **source()**
```{r}
source("http://tinyurl.com/rescale-R")
```

Test the rescale() function with some examples
```{r}
rescale(1:5)
```

```{r}
# rescale(c(1:5, "string"))
```

## Handling Errors and Unexpected Inputs

You can use **warning()** or **stop()** to warn users when using a function
warning() will continue to execute the function
stop() will stop the action of the function

the *call.* argument can display the line of code that produced the error 
Here is the example of the error handling of rescale2() function when encountering a non numeric value
```{r}
# rescale2(c(1:5,"string"))
```




Using the **is.numeric** function
```{r}
is.numeric(1:5)
```
## Practive writing a new function
```{r}
# Lets define an example x and y
x <- c( 1, 2, NA, 3, NA) 
y <- c(NA,3,NA,3, 4)
```

```{r}
is.na(x)
is.na(y)
# Comparing two vecotrs, or and
is.na(x) | is.na(y)
is.na(x) & is.na(y)
# This will show you the position of the missing value in a vector
which(is.na(x) & is.na(y))
```

Now we take our wokring snippet and turn it in to a function
```{r}
both_na <- function (x,y) {
  # Check for NA elements in both input vectors
  sum(is.na(x) & is.na(y))
}
```

Running the *both_na* function
```{r}
both_na(x,y)
```

Let's try to break the *both_na* function
```{r}
x <-  c(NA, NA, NA)
y1 <- c( 1, NA, NA)
y2 <- c( 1, NA, NA, NA)

both_na(x,y2)
# This will produce an error and an answer. The answer will be based on recycling of the value
```

# Installing a bioconductor package
```{r}
# source("http://bioconductor.org/biocLite.R")
# biocLite()
# biocLite("GenomicFeatures")
```



